Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-13 csf-2017/paper.tex
Took care of all the todos. Added two new ones.

We assume a finite set of register names $\RegName$. We define register files $\reg$ as
functions mapping register names to words and memories as functions that map
addresses to words. The state of the entire machine is represented as a
configuration that is either a running $\Phi \in \ExecConfs$ containing a memory
and a register file, or a failed or halted state, where the latter keeps hold
of the final state of memory. We also define memory segments $\ms$:
parts of memory represented as partial functions mapping addresses to words. \lau{we don't really make the distinction between memory and memory segments in the rest of the paper, so maybe we should leave this out?}

@@

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their current address and range
cannot be modified. They can only be used to jump to, but when that happens,
their permission is updated to $\exec$. This special behavior makes them
suitable for representing a kind of closures: an opaque package containing a
piece of code together with local encapsulated state. The idea is that one can
build an enter capability $c = ((\entry,\gl),\start,\addrend,\addr)$ where the
range $[\start,\addr-1]$ is used for storing the closure's local state (data or
capabilities) and the range $[\addr,\addrend]$ can be used for storing
instructions. The instructions and data are opaque to an adversary holding $c$
but when $c$ jumped to, the data becomes available to the closure's code through
the updated version of $c$ that is now in the $\pcreg$\footnote{Note that the
  capability in the $\pcreg$-register can be used to load from.}. \lau{As we
  talked about what is described here, is not how we make closures in the TR, so
  we should consider whether we should refrain from using that word here. }

@@

Explain our model of linking
\dominique{not sure what we want to say about it here...}

Explain malloc
\dominique{not sure what we want to say about it here either...}

@@

The basic idea of our approach is quite simple: we stick to a single, rather
standard, C stack and register-passed return pointers, much like a standard C
ABI\lau{This sentence does not compute}. However, to prevent various ways of
misusing this basic scheme, we put local capabilities to work and take a number
of not-always-obvious safety measures. In the next paragraphs, we will explain
the issues to be considered in all the relevant situations: when (1) starting
our program, (2) Returning to the adversary, (3) invoking the adversary, (4)
returning from the adversary, (5) invoking an adversary callback and (6) having
a callback invoked by the adversary.

\dominique{Add a summary table somewhere?}

@@

Already important is the fact that the return pointer must be provided to the
adversary as a local capability. If not, the adversary would be able to store
away the return pointer in a global data structure, and they would be able to jump
to it in circumstances where this should not be possible: for example, they may
store the return pointer, legally jump to it a first time, wait to be invoked
again and then jump to the old return pointer a second time, rather than the new
return pointer received for the new invocation. Similarly, they may store the
return pointer, invoke a function in our code, wait for us to invoke him again
and then jump to the stored, old return pointer rather than the new one he
received upon the second invocation. \lau{If we need space, then maybe one of the examples is sufficient.}
\lau{In the examples, we refer to the local return pointer as
  ``protected return pointer''. We should either introduce that here
  or use a different formulation in the examples. }

@@

Our answer is that the adversary should provide it to us, much like we provide
them with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
pointer and that its current address points just below its start address.
Without these checks, err...
\dominique{is this needed for security or just for convenience?}
\lau{Say we are called with a stack capability that is global and $\rwx$ as well as an integer as a non-local return pointer (this is fair game as the adversary has to follow no rules). Now you push the l(a)unch codes to what you think is the stack and \texttt{scall} the callback. At this point \texttt{scall} does its thing, but because it never attempts to store anything local it will succeed and jump to the adversary! The adversary can now look at what you thought was the stack and eat your lunch or fire the missiles as they prefer. It is not sufficient to require the stack to be just local because there may be a global equivalent. We know this is not the case for the rwlx capability because we have a system wide assumption that all rwlx capabilities are local.
Finally the adjustment of the pointer in \texttt{prepstck} (or whatever I called it) is a convenience/ space efficiency thing purely.}

TODO:
\begin{itemize}
\item Reusable macro-instructions for stack and return pointer management
\begin{itemize}
\item taking care of the stuff above
\end{itemize}
\end{itemize}

\lars{maybe we can present the examples informally before the logical relation and then return to them after the logical relation (to present the formal properties) ?}

@@

\subsection{Worlds}
\todo[inline]{Find a good place to describe monotonicity of
  interpretation function.}
\begin{figure}[htbp]
  \centering
  \includegraphics{w11}
  \caption{Illustration of the relation between local and global capabilities and temporary and permanent regions. The table represents a memory and the colored fields are regions governing parts of the memory. Roughly speaking, global capabilities should never depend on memory governed by temporary regions.}
  \label{fig:cap-world}
\end{figure}
A world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, which describes the protocol governing
the evolution of the memory described by the region, similarly to what
has been used in logical relations for high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. 
The protocols imposed on memory by permanent regions stay in place
indefinitely. This means that any capability, local or global, can
depend on these protocols. The protocols imposed on memory by temporary
regions can be revoked at certain points during execution,
corresponding to the points when local capabilities can be
revoked. This means that local capabilities safely can depend on the
protocols imposed on memory by temporary regions, whereas global
capabilities cannot, since a global capability may stay around
throughout the execution of a program, but a temporary region may be
revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

@@

The condition for capabilities with read permission is
$\readCond{}$. It ensures that a read capability can only be used to
read capability safe words, by requiring that the region that governs
the part of memory that the capability grants authority over enforces
that all the words are in the value relation. Generally speaking,
global capabilities should only depend on permanent regions. Therefore
we use the function $\var{localityReg}(g,W)$, which projects out all
active regions when $g$ is local, but only the permanent regions when
$g$ is global.
\todo[inline]{Do we want to mention the standard regions here? (They
  are not mentioned anywhere)}

@@

\section{Examples}
\label{sec:examples}
In this section, we demonstrate how our formal model of capability
safety allows us to prove local state encapsulation and control-flow
integrity properties for challenging program examples. The security
measures of Section~\ref{sec:stack-and-return-pointer} are deployed to
ensure these properties. Since we are dealing with assembly language,
there are many details to the formal treatment, and therefore we
necessarily have to omit some details in the lemmas we state here.
The interested reader can find all the technical details in the
techincal report~\cite{XXX}.

\subsection{Encapsulation of local state}
The capability machine provides encapsulation of local state. The
programs \texttt{f1} in \figurename~\ref{fig:prog-f1} and \texttt{f2}
\figurename~\ref{fig:prog-f2} demonstrate this.  The two programs
\texttt{f1} and \texttt{f2} are very similar. They both store some
local state, call an untrusted piece of code ($\adv$), and then tests
whether the local state is unchanged. They differ in the way they do
this. Program \texttt{f1} assumes it has access to a stack capability
and uses the stack to store its local state. It also uses the
stack-based calling convention \texttt{scall} to call the untrusted code. On the
other hand, \texttt{f2} allocates a new piece of memory for its local
state and uses an activation-record based calling convention
(described in the technical report) to run the adversarial code.

@@

\lau{in the following paragraph, we could make a reference to
  \textbf{invoking the adversary} from
  section~\ref{sec:stack-and-return-pointer} where the problem is
  first introduced. In fact, this paragraph is a reiteration of the
  problem in the aformentioned section.}
In \texttt{f3} there are two calls to an adversary and in order for
the assertion in the middle to succeed, they need to be
well-bracketed. If the adversary is able to store the protected return
pointer from the first call, so it is available for the second call, then it can be used
in a context where \texttt{f3} has $2$ on top of its stack which will cause the
assertion to fail. The protected return pointer is passed as a local
capability, so the only means the adversary has to store it is the
stack capability. The stack capability itself is a local capability,
so the adversary can only store it on the stack. This means that after
the adversary jumps to the protected return pointer, \texttt{f3} can
revoke it (by erasing it which happens as part of the
\texttt{scall}). In other words, the adversary cannot store the
protected return pointer for the second call, and it is possible to
prove that the assertion never fails.
\begin{figure}[htbp]
  \centering
  \begin{lstlisting}
f3: push 1
    fetch $r_1$ $\adv$
    scall $r_1$($[]$,$[r_1]$)
    pop $r_2$
    assert $r_2$ 1
    push 2
    scall $r_1$($[]$,$[]$)
    halt
\end{lstlisting}
  \caption{Simple program that relies on control-flow integrity.}
  \label{fig:prog-f3}
\end{figure}
\lau{Add something about the correctness lemma? Not sure if I have
  anything important to say.}
\begin{lemma}
  \label{lem:correctness-f3}
  Making similar assumptions about capabilities and linking as in
  Lemma~\ref{lem:correctness-f1}
  if $(\reg\update{\pcreg}{c_{f3}}\update{r_\stk}{c_\stk},m) \step^*
  (\halted,m')$, then $m'(\flag) = 0$.
\end{lemma}

@@

\begin{figure}[htbp]
  \centering
  \begin{lstlisting}
g1:  malloc $r_2$ 1
     store $r_2$ 0
     move $\pcreg$ $r_3$
     lea $r_3$ $\var{offset}$
     crtcls $[(x, r_2)]$ $r_3$
     rclear $\RegName \setminus \{\pcreg,r_0,r_1 \}$
     jmp $r_0$
f4:  reqglob $r_1$
     prepstack $r_\stk$
     store $x$ 0
     scall $r_1$($[]$,$[r_0,r_1,r_{\var{env}}]$)
     store $x$ 1
     scall $r_1$($[]$,$[r_0,r_{\var{env}}]$)
     load $r_1$ $x$
     assert $r_1$ 1
     mclear $r_\stk$
     rclear $\RegName \setminus \{r_0,\pcreg\}$
     jmp $r_0$
\end{lstlisting}
  \caption{A program that relies on well-bracketedness of
    \texttt{scall}s to function correctly. $\var{offset}$ is the
    offset to $f4$.}
  \label{fig:prog-g1}
\end{figure}

To illustrate how subtle this program is, consider one way an
adversary may try to make the assertion fail. In the second callback
an adversary can get to the first callback by calling the closure one
more time. If the adversary can save the protected return pointer in
the second callback and use it in the first, then the assertion will
fail because $x$ is $0$ before the first callback. The protected return
pointer cannot be transfered between the callbacks, so this does not
work.

>>>>

We assume a finite set of register names $\RegName$. We define register files $\reg$ as
functions mapping register names to words and memories as functions that map
addresses to words. The state of the entire machine is represented as a
configuration that is either a running $\Phi \in \ExecConfs$ containing a memory
and a register file, or a failed or halted state, where the latter keeps hold
of the final state of memory. 
% We also define memory segments $\ms$:
% parts of memory represented as partial functions mapping addresses to words.
%\lau{we don't really make the distinction between memory and memory segments in the rest of the paper, so maybe we should leave this out?}

@@

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their current address and range
cannot be modified. They can only be used to jump to, but when that happens,
their permission is updated to $\exec$. This special behavior makes them
suitable for representing a kind of closures: an opaque package containing a
piece of code together with local encapsulated state. The idea is that one can
build an enter capability $c = ((\entry,\gl),\start,\addrend,\addr)$ where the
range $[\start,\addr-1]$ is used for storing the closure's local state (data or
capabilities) and the range $[\addr,\addrend]$ can be used for storing
instructions. The instructions and data are opaque to an adversary holding $c$
but when $c$ is jumped to, the data becomes available to the closure's code through
the updated version of $c$ that is now in the $\pcreg$\footnote{Note that the
  capability in the $\pcreg$-register can be used to load from.}.
% \lau{As we
%   talked about what is described here, is not how we make closures in the TR, so
%   we should consider whether we should refrain from using that word here. }

@@

In order to have a reasonably realistic system, we have created a
simple model of linking, so programs can have access to other programs
through a linking table. We also assume malloc to be part of the
trusted computing base satisfying a certain specification. Malloc and
the model linking is described to the degree necessary, in the next
section, but we refer to the technical appendix~\ref{XXX} for further
details.

@@

The basic idea of our approach is quite simple: we stick to a single, rather
standard, C stack and register-passed return pointers, much like a standard C
ABI. However, to prevent various ways of
misusing this basic scheme, we put local capabilities to work and take a number
of not-always-obvious safety measures. In the next paragraphs, we will explain
the issues to be considered in all the relevant situations: when (1) starting
our program, (2) Returning to the adversary, (3) invoking the adversary, (4)
returning from the adversary, (5) invoking an adversary callback and (6) having
a callback invoked by the adversary.

% \dominique{Add a summary table somewhere?}
% \lau{It seems like we want have the space for this}.

@@

Already important is the fact that the return pointer must be provided to the
adversary as a local capability. If not, the adversary would be able to store
away the return pointer in a global data structure, and they would be able to jump
to it in circumstances where this should not be possible: for example, they may
store the return pointer, legally jump to it a first time, wait to be invoked
again and then jump to the old return pointer a second time, rather than the new
return pointer received for the new invocation. Similarly, they may store the
return pointer, invoke a function in our code, wait for us to invoke him again
and then jump to the stored, old return pointer rather than the new one they
received upon the second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of Section~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}

@@

Our answer is that the adversary should provide it to us, much like we provide
them with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
capability and make sure its current address points just below its start address.
Without this check, an adversary could potentially get control over
our local stack frame during a callback by passing us a local $\rwx$
capability to a global data structure (i.e., there exists a global
capability for it) as the stack capability and a global
return capability. If our local state contains no local capabilities,
then following our calling convention, the callback would not fail and
the adversary could use a stored capability for the global data
structure to access our local state. 
To prevent this from happening, we need to make sure the stack
capability is local and there exists no global capability for that
memory segment. By making sure the capability has permission $\rwlx$,
we rely on the system wide assumption that there are only local
$\rwlx$ capabilities.

\textbf{Reusable macro instructions} With the high-level approach in
mind, we define the following macro instructions: 
\texttt{fetch $r$ $\var{name}$} fetches the capability related to
$\var{name}$ from the linking table and stores it in register
$r$. 
\texttt{push $r$}/\texttt{pop $r$} works as one would expect. All
macros that uses the stack assumes a stack capability in
register $r_\stk$. 
\texttt{prepstk $r$} prepares the stack by checking it has permission
$\rwlx$ and making it conform to our stack discipline. 
\texttt{scall  $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}
Jumps to the capability in register $r$ in a safe manner. That is it
pushes local state, stack capability and in code return capability to
the stack, sets up the necessary on stack return code and creates a
return capability, restricts stack capability, clears the unused part
of the stack, clears the necessary registers and jumpsto $r$. Upon
return, the private state is restored.
\texttt{mclear $r$} clears all the memory the capability in register
$r$ has authority over.
\texttt{rclear $\var{regSet}$} clears all the registers in $\var{regSet}$
\texttt{reqglob $r$} checks whether the word in register $r$ is a
global capability.
\texttt{crtcls $\overline{(x_i,r_i)}$ $r$} allocates a closure pair
where the capability in register $r$ is used to point to the program
and a new environment is allocated where the contents of $\overline{r_i}$ is
stored. In the program referred to by $r$, an implicit fetch
happens when an instruction refers to $x_i$.

In Section~\ref{sec:examples}, we use the above macros and some
more. The technical appendix~\ref{XXX} contains detailed descriptions
of all the macros.

%\lars{maybe we can present the examples informally before the logical
%relation and then return to them after the logical relation (to
%present the formal properties) ?}
% \lau{not enough space to do this.}

@@

\subsection{Worlds}
%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}
\begin{figure}[htbp]
  \centering
  \includegraphics{w11}
  \caption{Illustration of the relation between local and global capabilities and temporary and permanent regions. The table represents a memory and the colored fields are regions governing parts of the memory. Roughly speaking, global capabilities should never depend on memory governed by temporary regions.}
  \label{fig:cap-world}
\end{figure}
A world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, which describes the protocol governing
the evolution of the memory described by the region, similarly to what
has been used in logical relations for high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. 
The protocols imposed on memory by permanent regions stay in place
indefinitely. This means that any capability, local or global, can
depend on these protocols. The protocols imposed on memory by temporary
regions can be revoked at certain points during execution,
corresponding to the points when local capabilities can be
revoked. This means that local capabilities safely can depend on the
protocols imposed on memory by temporary regions, whereas global
capabilities cannot, since a global capability may stay around
throughout the execution of a program, but a temporary region may be
revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

@@

The condition for capabilities with read permission is
$\readCond{}$. It ensures that a read capability can only be used to
read capability safe words, by requiring that the region that governs
the part of memory that the capability grants authority over enforces
that all the words are in the value relation. Generally speaking,
global capabilities should only depend on permanent regions. Therefore
we use the function $\var{localityReg}(g,W)$, which projects out all
active regions when $g$ is local, but only the permanent regions when
$g$ is global.
The read condition uses one of two standard regions found
in~\ref{fig:standard-regions}. $\iota^\pwl_{\start,\addrend}$ requires
all the words in the range $[\start,\addrend]$ to be capability
safe. $\iota^\nwl_{\start,\addrend}$ requires the words to be
non-local and capability safe.

@@

\section{Examples}
\label{sec:examples}
In this section, we demonstrate how our formal model of capability
safety allows us to prove local state encapsulation and control-flow
integrity properties for challenging program examples. The security
measures of Section~\ref{sec:stack-and-return-pointer} are deployed to
ensure these properties. Since we are dealing with assembly language,
there are many details to the formal treatment, and therefore we
necessarily have to omit some details in the lemmas we state here.
The interested reader can find all the technical details in the
techincal appendix~\cite{XXX}.

\subsection{Encapsulation of local state}
The capability machine provides encapsulation of local state. The
programs \texttt{f1} in \figurename~\ref{fig:prog-f1} and \texttt{f2}
\figurename~\ref{fig:prog-f2} demonstrate this.  The two programs
\texttt{f1} and \texttt{f2} are very similar. They both store some
local state, call an untrusted piece of code ($\adv$), and then tests
whether the local state is unchanged. They differ in the way they do
this. Program \texttt{f1} assumes it has access to a stack capability
and uses the stack to store its local state. It also uses the
stack-based calling convention \texttt{scall} to call the untrusted code. On the
other hand, \texttt{f2} allocates a new piece of memory for its local
state and uses an activation-record based calling convention
(described in the technical appendix) to run the adversarial code.

@@

\lau{in the following paragraph, we could make a reference to
  \textbf{invoking the adversary} from
  section~\ref{sec:stack-and-return-pointer} where the problem is
  first introduced. In fact, this paragraph is a reiteration of the
  problem in the aformentioned section.}
In \texttt{f3} there are two calls to an adversary and in order for
the assertion in the middle to succeed, they need to be
well-bracketed. If the adversary is able to store the return
pointer from the first call, so it is available for the second call, then it can be used
in a context where \texttt{f3} has $2$ on top of its stack which will cause the
assertion to fail. The return pointer is passed as a local
capability, so the only means the adversary has to store it is the
stack capability. The stack capability itself is a local capability,
so the adversary can only store it on the stack. This means that after
the adversary jumps to the return pointer, \texttt{f3} can
revoke it (by erasing it which happens as part of the
\texttt{scall}). In other words, the adversary cannot store the
return pointer for the second call, and it is possible to
prove that the assertion never fails.
\begin{figure}[htbp]
  \centering
  \begin{lstlisting}
f3: push 1
    fetch $r_1$ $\adv$
    scall $r_1$($[]$,$[r_1]$)
    pop $r_2$
    assert $r_2$ 1
    push 2
    scall $r_1$($[]$,$[]$)
    halt
\end{lstlisting}
  \caption{Simple program that relies on control-flow integrity.}
  \label{fig:prog-f3}
\end{figure}
\lau{Add something about the correctness lemma? Not sure if I have
  anything important to say.}
\begin{lemma}
  \label{lem:correctness-f3}
  Making similar assumptions about capabilities and linking as in
  Lemma~\ref{lem:correctness-f1}
  if $(\reg\update{\pcreg}{c_{f3}}\update{r_\stk}{c_\stk},m) \step^*
  (\halted,m')$, then $m'(\flag) = 0$.
\end{lemma}

@@

\begin{figure}[htbp]
  \centering
  \begin{lstlisting}
g1:  malloc $r_2$ 1
     store $r_2$ 0
     move $\pcreg$ $r_3$
     lea $r_3$ $\var{offset}$
     crtcls $[(x, r_2)]$ $r_3$
     rclear $\RegName \setminus \{\pcreg,r_0,r_1 \}$
     jmp $r_0$
f4:  reqglob $r_1$
     prepstk $r_\stk$
     store $x$ 0
     scall $r_1$($[]$,$[r_0,r_1,r_{\var{env}}]$)
     store $x$ 1
     scall $r_1$($[]$,$[r_0,r_{\var{env}}]$)
     load $r_1$ $x$
     assert $r_1$ 1
     mclear $r_\stk$
     rclear $\RegName \setminus \{r_0,\pcreg\}$
     jmp $r_0$
\end{lstlisting}
  \caption{A program that relies on well-bracketedness of
    \texttt{scall}s to function correctly. $\var{offset}$ is the
    offset to $f4$.}
  \label{fig:prog-g1}
\end{figure}

To illustrate how subtle this program is, consider one way an
adversary may try to make the assertion fail. In the second callback
an adversary can get to the first callback by calling the closure one
more time. If the adversary can save the return pointer in
the second callback and use it in the first, then the assertion will
fail because $x$ is $0$ before the first callback. The return
pointer cannot be transfered between the callbacks, so this does not
work.
